An Observational Study to COVID-19 and 2020 US Presidential Election: Counties with High Deaths per Case Rate Showed Diminished Support For Trump

### 2023-12-15 outputs/paper/paper.qmd
Update the introduction (First Draft)

---
title: "Causal Inference in the 2020 US Federal Election: While COVID-19 Lowered Trump's Support, Loss of Confidence Among Wealthier Voters Seems More Decisive"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss.git"
date: "today"
date-format: "long"
abstract: "This study used the election data from MIT EDSL and COVID-19 data from JHU CSSE, to investigate the causal inference between COVID-19 and Donald Trump's loss during the 2020 US Federal Election. The main methodology used in this paper is Propensity Score Matching with testing different treatments. The primary finding is that the counties with high death per case rate seems to vote less for Trump with approximate treatment effect -0.01. Besides, it seems that the interaction effect with income seems to be more critical: the medium-high income levels counties seems to vote at maximum 3% less for Trump compared to those low income counties. Furthermore, this paper also conduct counterfactual analysis and indicate the Trump would pre-elect if there was no COVID. Furture analysis should focus more on the choice of covariates to predict propensity scores."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(dplyr)
library(knitr)
library(ggplot2)
library(maps)
library(mapdata)
library(RColorBrewer)
library(gridExtra)
library(tools)
```

@@

@@

@@

In addition to that, the study also uses the data regarding COVID-19 from the Center for Systems Science and Engineering at Johns Hopkins University

In this paper, the data used mainly consists of the voting patterns for the 2020 US Federal Election, COVID infection rate and socioeconomic variables in each county of the US. The voting data is taken from the MIT Election Data Science Lab, including voting for each party at the county level at each Federal Election from 2000 to 2020. The COVID data is from the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University. The socioeconomic data is taken from the American Community Survey.

## MIT Election Data Science Lab

The MIT Election Data Science Lab is a lab at MIT that collects and analyzes election data to support advances in election science. Their collected data is not limited to the presidential election results but also covers the midterm or US Senate results at the state or county level.

The data abstracted from MITEDSL in this paper is the "County Presidential Election Returns 2000-2020." It contains the number of votes for each party and the total votes at each US county from 2000 to the 2020 US Federal Election. In this paper, I will only use the data for the 2020 US Election and focus on the counties on the mainland of the US. The @tbl-election summarizes the essential variables.

```{r}
#| echo: false
#| message: false
#| label: tbl-election
#| tbl-cap: The summary of voting patterns in 2020 US Federal Presidential Election
# election_data %>% 
#   group_by(party, candidate) %>% 
#   summarise(total = sum(votes),
#             mean = round(mean(pct_vote), 2),
#             median = round(median(pct_vote), 2)) %>%
#   arrange(desc(total)) %>% 
#   rename(`Party` = party,
#          `Candidate` = candidate,
#          `Total Votes` = total,
#          `Mean Pct Vote` = mean,
#          `Median Pct Vote` = median) %>% 
#   kable()
```

The @tbl-election summarizes the voting patterns for each party during the 2020 U.S. Presidential Election. Undoubtedly, the two most popular parties are the Democrats and the Republicans. However, even though Republican has almost double the mean percent vote than Democrat, the total votes for Republican is less than for Democrat. This pattern may indicate that more people living in high-population states, such as California, voted for Biden. This finding aligns with the fact that Biden defeated Trump in 2020. I will explain more later.

## Center for Systems Science and Engineering at JHU

The Center for Systems Science and Engineering at JHU is at the Department of Civils and Engineering, which collects local, national, and global multidimensional data, including medicine, health care, disaster response, etc. During the pandemic, they collected the U.S. and international COVID cases and deaths and reported them on their GitHub. Their data is summarized by daily reports ranging from April 12, 2020, to March 9, 2023.

To illustrate the impact of COVID-19 on the 2020 Election to the greatest extent, the COVID data used in this paper will be the daily report on November 3, 2020, which is Election Day \[citation\]. This report contains the aggregated cases and deaths for each country and counties in the U.S.

## American Community Survey (ACS)

The American Community Survey is the survey conducted by the U.S. Census Bureau. The survey contains a variety of socio-economic variables for each county. Although there are different data tables from ACS, I will use the 2020 five-year estimate of DP02, DP03, and DP05.

These three tables cover the social, economic and demographic characteristics of each county in the U.S. By referring to my previous research, where I found the variables with the highest correlations to the COVID mortality rate, I will use the same variables in this paper. That said, I will extract the data regarding the educational attainment from DP02, especially the proportion of people having at least a bachelor's degree. Regarding the economy, I will take the proportion of people with private health insurance and the mean household income. Lastly, regarding the demographic factors, I will use the total population, the ratios of children and individuals above 85. I will also take the White and Black people percentage as well.

## Data Cleaning

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(case = mean(cases),
            deaths = mean(deaths),
            inf = mean(infrate),
            dpc = mean(dpc),
            income = mean(mean_household_income)) %>% 
  rename(`Winning Party` = winning_party,
         `Cases` = case,
         `Deaths` = deaths,
         `Infection Rate` = inf,
         `Income` = income) %>% 
  kable()
```

Combining the three data sources, @tbl-covid shows the COVID cases and deaths and the infection and mortality rate for the county voting for the Democrat and Republican parties, respectively. Surprisingly, the average cases and fatalities of COVID-19 in the counties voting for the Democrats are significantly higher than that of the Republican party. However, the values of infection and mortality rate do not show the same dramatic difference. It also emphasizes that the counties (states) voting for the Democrat are population intensive, and counties with more residence generally have higher mean household income levels. We can observe this from the average income where the "Democratic" counties have about \$13k mean income more elevated than the "Republic" ones.

@@

```{r}
binned_data <- data %>%
  mutate(dpc_pctile = ntile(dpc, 100)) %>%  # Adjust 'breaks' to change number of bins
  group_by(dpc_pctile) %>%
  summarize(mean_rep = mean(pct_vote_rep, na.rm = TRUE),
            mean_demo = mean(pct_vote_demo, na.rm = TRUE)) %>% 
  mutate(bin = 1:n())

binned_data2 <- data %>%
  mutate(dpc_pctile = ntile(mean_household_income, 100)) %>%  # Adjust 'breaks' to change number of bins
  group_by(dpc_pctile) %>%
  summarize(mean_rep = mean(pct_vote_rep, na.rm = TRUE),
            mean_demo = mean(pct_vote_demo, na.rm = TRUE)) %>% 
  mutate(bin = 1:n(),
         sum = mean_rep + mean_demo)

# Plotting
ggplot(binned_data, aes(x = bin, y = mean_rep)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm",formula = y ~ poly(x, 4), se=FALSE) +
  theme_minimal() +
  labs(x = "Binned mortality per case ", y = "Mean vote for Trump",
       title = "Votes for Trump vs. MPC")
ggplot(binned_data, aes(x = bin, y = mean_demo)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm", formula = y ~ poly(x, 4),se=FALSE) +
  theme_minimal() +
  labs(x = "Binned mortality per case ", y = "Mean vote for Biden", 
       title = "Votes for Biden vs. MPC")

ggplot(binned_data2, aes(x = bin, y = mean_rep)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se=FALSE) +
  theme_minimal() +
  labs(x = "Binned income levels", y = "Mean vote for Trump", 
       title = "Votes for Trump vs. Income")
ggplot(binned_data2, aes(x = bin, y = mean_demo)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se=FALSE) +
  theme_minimal() +
  labs(x = "Binned income levels", y = "Mean vote for Biden",
       title = "Votes for Biden vs. Income")
```

```{r}
#| echo: false
#| message: false
#| fig.width: 10 
#| fig.height: 10
#| label: fig-map
#| fig-cap: The ratio of votes for the Republican and the infection rate per 100k in each county

p1 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = pct_vote_rep), color = "black") +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "grey") +
  labs(title = "The relative share of votes between Republican and Democrat party",
       fill = 'Share of Republican') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Second plot
p2 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = dpc), color = "black") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "grey",
                      limits = c(0, 5000), oob = scales::squish) +
  labs(title = "The distribution of death per case rate by 100K",
       fill = 'Death per cases by 100k') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

p3 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = mean_household_income), color = "black") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "grey",
                      limits = c(60000, 120000), oob = scales::squish) +
  labs(title = "The distribution of mean household income",
       fill = 'Mean Hosehold Income') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Arrange the plots in a 2x1 layout
grid.arrange(p1, p3, p2, ncol = 1)
```

```{r}
#| eval: false
library(sp)
library(geojsonio)
library(ggplot2)
library(dplyr)
library(broom)
hex_county <- geojson_read("us_county_hexgrid.geojson", what = "sp")

# Reformat the data as needed (depending on your data structure)
hex_county@data <- hex_county@data %>% 
  mutate(adjusted_field = gsub("some_pattern", "replacement", original_field))

# Fortify the data for ggplot
hex_county_fortified <- tidy(hex_county, region = "desired_region_field")

# Plot the hexagon map
ggplot() +
  geom_polygon(data = hex_county_fortified, aes(x = long, y = lat, group = group), fill="#a1dab4", color="#f7f7f7") +
  geom_text(data = hex_county_fortified, aes(x = long, y = lat, label = label_field), size = 3) +
  theme_void() +
  coord_map()
```

## Data Limitation

Although the merged data covers the voting patterns, socio-economic characteristics and COVID-19 infection rate by county, some counties are missing. From the above two maps, the counties in the middle west show grey, indicating that the data for these counties are absent. This absence of data may increase bias and influence the accuracy of data analyses.

@@

This paper aims to make a causal inference between the 2020 voting patterns and COVID-19 to find whether Trump can contribute to his loss of COVID-19. To see the causal effect in the observational study, I will implement the methods of propensity scores to find the change in votes for Trump due to the pandemic. Based on the results, I will also conduct a counterfactual analysis to see what would have happened if there had been no COVID-19,

First and foremost, we need to define the treatment. The treatment in an experiment will be artificially imposed on specific groups of people to see the effect of interested variables with and without this treatment. Therefore, the people with the treatment are called the treatment group, and the rest are called the control group. The treatment and control group settings are to test whether the imposed factor impacts the interested variable. However, since this is an observational study, it is unrealistic to set a factor manually. Therefore, to find the causal effect of COVID-19 on Trump's voting, I will choose the treatment to be whether a county has a high infection rate.

## Propensity Score

Propensity score is a technique that can help us to decide which county would be in the treatment group. That said, the propensity score in this study is the probability of a county having a high infection rate, given a set of socioeconomic variables. Using the propensity score can ensure that the two counties with similar propensity scores have the same distribution of the observed socioeconomic variables, such as income level, even though they are in different groups. This is fantastic, as we can have many pairs from the treatment of the control group with similar covariates by fixing propensity scores. The key drawback is that we can not estimate the causal effect from observational studies.

>>>>

---
title: "An Observational Study to COVID-19 and 2020 US Presidential Election: Counties with High Deaths per Case Rate Showed Diminished Support For Trump"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss.git"
date: "today"
date-format: "long"
abstract: "This study investigate the causal inference between COVID-19 and Donald Trump's loss during the 2020 US Federal Election using the data from MIT Election Data Science Lab and Johns Hopkins University CSSE. The main methodology used in this paper is Propensity Score Matching with with an exploratory analyzing the optimal treatment initially. The key finding suggest that the counties with a death per case rate exceeding the 0.4 quantile threshold shows a reduced voting preference for Trump. Additionally, this paper conducts a counterfactual analysis based on the treatment effect, indicating that Trump might have be secured to re-elect if the disparities in death per case rates were addressed. Future research should aim to refine the propensity score calculations by incorporating additional variables, such as the winning party in each county during the 2016 election."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(dplyr)
library(knitr)
library(ggplot2)
library(maps)
library(mapdata)
library(RColorBrewer)
library(gridExtra)
library(tools)
library(dplyr)
library(ggplot2)
```

@@

At the beginning of 2020, US reported the first COVID-19 cases. This number grows to ten thousand within two months on March 19 and increase dramatically to 100K on March 27 [citation]. At the same year, the U.S. Federal Presidential Election started off and Trump lost to Biden. As of Election Day, there was above 9 million cases and about 0.2 million deaths in US. When people talk about Trump's lost, people always connect it to the pandemic, saying that he "mishandled his greatest test" [citation] and 2016 Election was a historical accident [citation]. However, there is another voice that COVID is not the only or the decisive factor attributing to his lost, his limits as a politican and wasted his advantage on the economy are also important explaining why he lost. This paper will investigate whether the is a causal effect between COVID and Trump's lost. If there is, whether it is strong enough to beath Trump at 2020.

There are existing researches on this. Leaonardo et al. [citation] suggested that Trump could win if there was no COVID but it will be too naive to simply connect his lost to the COVID-19 infection rate. Their research highlighted that the COVID-19 had the most significant negative impacts on those urban areas where there was no stay-at-home orders, especially for those "swing" states. They also argue that the race diversity and education attainment are also important when analysing voting patterns for Trump. In addition to that, Marcus et al. [citation] suggested that the deaths per case is a more crucial indicator than infection rate when analysing the voting for Trump. However, they suggest the voting itself may not capture the impact of COVID as we do not know when did the voters make their decision, especially for early vote. Compared to these researches, Harold [citation] conduct read survey before and after the election to illustrate the impact of COVID pattern, they found that even though COVID has impacts on voting, it is not the dominant factor. Without COVID, Trump was also less likely to re-elect due to US highly polarized political landscape. Similarly, Shang et al. [citation] also conduct a survey and all the participants show natural or negative altitude. They suggest some social activity like "Black lives matter" may have impacts on voting. Furthermore, the

While all the above researches had investigated the correlation between COVID-19 and votings for Trump. Except conducting real surveys, the methods they used were to conduct a model and find the relevant variables in predicting the (change) vote for Trump. Then conduct counterfactual analysis to predict what will be the voting for Trump if there was no COVID or a certain amount reduce in COVID deaths. Instead using a similar way like those studies, this paper will introduce the method of Propensity Score Matching to find the treatment effect, to see whether those more impacted counties will have different voting patterns for Trump. However, since all counties experienced COVID-19, this paper will firstly find the optimal treatment groups. After that, I will also conduct a counterfactual analysis to see if Trump could re-elect.

There will be five parts in this paper. I will introduce the data used in this paper and present data suammries and visualization in [Data] part. After that, I will introduce the methods in this paper and the corresponding results in  [Methods] and [Results] part, respectively. Then the results will be interpreted and discussed in the [Discussion] part and conclude with limitations and drawbacks in [Conclusion] part..

@@

## Data sources

@@

In addition to that, the study also uses the data regarding COVID-19 from the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University. CSSE collects and reports the local, national, and global multidimensional data, including medicine, health care, disaster response, etc. During the pandemic, they collected the U.S. and international COVID cases and deaths and reported them on their GitHub, summarized by daily reports from April 12, 2020, to March 9, 2023. To access the impact of COVID on 2020 Election, this paper used the daily report on November 3, 2020 which is the election day of 2020 US Federal Election. The resulting data include the aggregates number of cases, deaths and recovers, as well as the incidence rate and case fatility ratio for each county by the election day.

Lastly, this paper also employ the socio-economic data from American Community Survey (ACS). ACS is a online open source database conducted by the U.S. Census Bureau, containing the various soci-economic factors in either geographic levels. The data sets extracted from ACS is 2020 five-year estimates of DP02, DP03 and DP05 which covers the social, economic and demographic characteristics in county level. Since there are thousands of variables, I will only use the variables which are found from my previous paper that are significant in predicting the COVID-19 mortality rate such as the mean household income, high-education attainment etc. The descriptions of all variables can be found on Table 1.

After having five data sets from three sources, they are merged into one big data by counties. Using the merged data, there are some new variables calculated by the exiting variables. In terms of election, I calculate the percentage of vote for each party (candidate) in both 2016 and 2020 Election, and calculate the corresponding change of percentage votes. I also create new variables indicating the winning party in both elections for each county. Additionally, to accurately compare the COVID cases and deaths across all counties, I transform the number of cases and deaths to infection and mortality rate by 10,000 citizens in each county. The final data now consists of 3107 rows with 36 columns. All the important variables are described in Table 1.

```{r}
#| tbl-cap: Descriptions of all important variables in the analyze data
#| label: tbl-descriptions
data.frame(var = c("Income Percentile", "High-Education Attainment",
                   "Private Insurance", "No Insurance", "White Population",
                   "Black Population", "Males Population", "Infection Rate",
                   "Mortality Rate", "Death per Case", "Votes for Democrat (2016)",
                   "Votes for Democrat (2020)", "Votes for Republican (2016)",
                   "Votes for Republican (2020)", "Change of Vote for Republican"),
           code = c("income_pctile", "prop_high_education", "private_insurance",
           "no_insurance", "white_pct", "black_pct", "males", "infrate",
           "mortrate", "dpc", "pct_vote_demo16", "pct_vote_demo", 
           "pct_vote_rep16", "pct_vote_rep", "change_vote_rep"),
           descriptions = c("The income percentile of each county",
                            "The proportion of residents in a county having a at least bachelor degree",
                            "The proportion of local residences having private insurance",
                            "The proportion of local residences without any health insurance",
                            "The proportion of White population",
                            "The proportion of Black population",
                            "The proportion of Males population",
                            "The COVID infection rate, calculated by number of case per 10,000 residences",
                            "The COVID mortality rate, calculated by number of death per 10,000 residences",
                            "The COVID death per 10,000 confirmed cases",
                            "The percentage of votes for the Democrat in 2016",
                            "The percentage of votes for the Democrat in 2020",
                            "The percentage of votes for the Republican in 2016",
                            "The percentage of votes for the Republican in 2020",
                            "The change of vote for the Republican from 2016 to 2020")) %>% 
  rename(`Variables` = var,
         `Coded name` = code,
         `Descriptions` = descriptions) %>% 
  kable(booktabs = TRUE)
```

## Data summaries and visualizations

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(n = n(),
            case = mean(cases),
            inf = mean(infrate),
            deaths = mean(deaths),
            mort = mean(mortrate),
            dpc = mean(dpc),
            income = mean(mean_household_income)) %>% 
  rename(`Winning Party` = winning_party,
         `Count` = n,
         `Cases` = case,
         `Infection Rate` = inf,
         `Deaths` = deaths,
         `Mortality Rate` = mort,
         `DPC` = dpc,
         `Income` = income) %>% 
  kable(booktabs = TRUE)
```

@tbl-covid summarize the COVID-19 impacts and income levels between the counties voting for Democrat and Republican. Even though the Republican won about five sixths counties, there is a stark contrast where counties supporting Biden had nearly ten times the average number of COVID-19 cases and deaths compared to those supporting Trump. Despite a similar infection and mortality rates between the two groups, there is still a notable higher death-to-case rate in counties voting for the Republican. Furthermore, these counties also has a lower mean household income than the "Democrat-win" counties. These patterns suggest a potential correlation between the extent of COVID-19 impact and voting behavior in the 2020, where areas more severely affected by COVID-19 were less likely to vote for the Trump, vice-versa. This aligns with mainstream media and intuitive expectations.

@@

```{r}
#| echo: false
#| message: false
#| fig.width: 8 
#| fig.height: 8
#| label: fig-map
#| fig-cap: The ratio of votes for the Republican and the infection rate per 100k in each county

p1 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = pct_vote_rep), color = "black") +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "grey") +
  labs(title = "The relative share of votes between Republican and Democrat party",
       fill = 'Share of Republican') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.9)

# Second plot
p2 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = dpc), color = "black") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "grey",
                      limits = c(0, 5000), oob = scales::squish) +
  labs(title = "The distribution of death per case rate by 100K",
       fill = 'Death per cases by 100k') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.9)

# Arrange the plots in a 2x1 layout
grid.arrange(p1, p2, ncol = 1)
```

We can validate our guess from @fig-map which compares the relative ratio of votes for the two parties and the death to cases rate in maps. From the maps, it is clearly more counties had preference for the Republican but the counties with higher death per case rate have a smaller ratio of vote for the Republican and these counties are usually richer ones. The most significant example can be California and New York. In contrast, the states with less impacted by COVID, such as Utah, vote more for Republican.

```{r}
#| fig-cap: The correlation between voting behaviors to income and DPC rate for the Republican and Democrat
#| label: fig-dpc-income

binned_data <- data %>%
  mutate(dpc_pctile = ntile(dpc, 100)) %>%  
  group_by(dpc_pctile) %>%
  summarize(mean_rep = mean(pct_vote_rep, na.rm = TRUE),
            mean_demo = mean(pct_vote_demo, na.rm = TRUE),
            mean_vote_rep = mean(vote_rep, na.rm = TRUE),
            mean_vote_demo = mean(vote_demo, na.rm = TRUE),
            mean_income = mean(mean_household_income))

binned_data2 <- data %>%
  mutate(income_pctile = ntile(mean_household_income, 100)) %>%  
  group_by(income_pctile) %>%
  summarize(mean_rep = mean(pct_vote_rep, na.rm = TRUE),
            mean_demo = mean(pct_vote_demo, na.rm = TRUE),
            mean_vote_rep = mean(vote_rep, na.rm = TRUE),
            mean_vote_demo = mean(vote_demo, na.rm = TRUE)) %>% 
  mutate(sum = mean_rep + mean_demo)

plot1 <- ggplot(binned_data, aes(x = dpc_pctile)) +
  geom_point(aes(y = mean_rep, colour = "Republican"), alpha = 0.25) +  
  geom_smooth(aes(y = mean_rep, colour = "Republican"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_point(aes(y = mean_demo, colour = "Democrat"), alpha = 0.25) + 
  geom_smooth(aes(y = mean_demo, colour = "Democrat"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_vline(xintercept = 25, 
             color = "black", linetype = "dashed", size = 0.6) +
  geom_vline(xintercept = 75, 
             color = "black", linetype = "dashed", size = 0.6) +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Party",  
                      labels = c("Democrat", "Republican")) +
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(x = "DPC Rate Percentiles", 
       y = "Votes Share (%)")

plot2 <- ggplot(binned_data, aes(x = dpc_pctile)) +
  geom_point(aes(y = mean_vote_rep, colour = "Republican"), alpha = 0.25) +  
  geom_smooth(aes(y = mean_vote_rep, colour = "Republican"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_point(aes(y = mean_vote_demo, colour = "Democrat"), alpha = 0.25) + 
  geom_smooth(aes(y = mean_vote_demo, colour = "Democrat"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_vline(xintercept = 40, 
             color = "black", linetype = "dashed", size = 0.6) +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Party",  
                      labels = c("Democrat", "Republican")) +
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(x = "DPC Rate Percentiles", 
       y = "Number of Votes")

pred <- predict(lm(mean_rep ~ poly(income_pctile, 3), data = binned_data2), 
                           newdata = binned_data2)

plot3 <- ggplot(binned_data2, aes(x = income_pctile)) +
  geom_point(aes(y = mean_rep, colour = "Republican"), alpha = 0.25) +  
  geom_smooth(aes(y = mean_rep, colour = "Republican"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_point(aes(y = mean_demo, colour = "Democrat"), alpha = 0.25) + 
  geom_smooth(aes(y = mean_demo, colour = "Democrat"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_vline(xintercept = which.max(pred), 
             color = "black", linetype = "dashed", size = 0.6) +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Party",  
                      labels = c("Democrat", "Republican")) +
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(x = "Income Percentiles", 
       y = "Votes Share (%)")

plot4 <- ggplot(binned_data2, aes(x = income_pctile)) +
  geom_point(aes(y = mean_vote_rep, colour = "Republican"), alpha = 0.25) +  
  geom_smooth(aes(y = mean_vote_rep, colour = "Republican"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_point(aes(y = mean_vote_demo, colour = "Democrat"), alpha = 0.25) + 
  geom_smooth(aes(y = mean_vote_demo, colour = "Democrat"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_vline(xintercept = 25, 
             color = "black", linetype = "dashed", size = 0.6) +
  geom_vline(xintercept = 78, 
             color = "black", linetype = "dashed", size = 0.6) +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Party",  
                      labels = c("Democrat", "Republican")) +
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(x = "Income Percentiles", 
       y = "Number of Votes")

library(cowplot)
legend <- get_legend(plot1 + theme(legend.position = "bottom"))
combined_plot <- plot_grid(plot1, plot2, plot3, plot4, ncol = 2)
combined_plot_with_legend <- plot_grid(combined_plot, legend, ncol = 1, rel_heights = c(1, 0.1))

print(combined_plot_with_legend)
```

Given the insights from @tbl-covid and @fig-map, @fig-dpc-income further examine the impact of COVID and income levels on the voting behaviors for the two parties. @fig-dpc-income shows the correlation between the votes of the two parties to the death per case rate and income levels. The two graphs on the left indicates the share of votes whereas the right two graphs indicate the number of votes. The two parties are represented by different colors. In terms of share of votes, in general, there is a positive relationship between the DPC and votes for the Democrat but a negative one for the Republican. However, the counties with medium value of DPC is less sensitive to the voting for the two parties, compared to the "tail" counties which their support change significantly as DPC increases or decreases. Meanwhile, there is a quadratic patterns between the vote for the two parties and income, it seems that the Republican is in favor of low income counties, but as income increase, more counties vote for Bide not Trump.

In addition, even though the average share of votes for Republican is always higher than 0.5, it does not necessarily mean that the Democrat lost in all counties. The reason behind this is, as mentioned in @tbl-covid, the counties which the Republican won is about five times as higher as the Democrat. Since Republican has more counties, it definitely has a higher average share of votes than Democrat. This pattern also indicate that Democrat usually won the densely populated, urban areas but Republican won in more sparsely populated, rural areas \[citation\]. The high population density areas has more votes and explains why Biden beat Trump. We can verify this from the number of votes for the two parties, where the number of votes for Biden exceeds Trump as DPC and income level increases. This pattern is especially significant for the top 50% richest counties and the counties with DPC larger than 75 quantile.

```{r}
data_population <- data %>% 
  mutate(pop_pctile = ntile(total_population, 100)) %>% 
  group_by(pop_pctile) %>% 
  summarise(county_demo = sum(winning_party == "Democrat")/n(),
            county_rep = sum(winning_party == "Republican")/n())

data_population %>% 
  ggplot(aes(x = pop_pctile)) +
  geom_point(aes(y = county_rep, colour = "Republican"), alpha = 0.25) +  
  geom_smooth(aes(y = county_rep, colour = "Republican"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_point(aes(y = county_demo, colour = "Democrat"), alpha = 0.25) + 
  geom_smooth(aes(y = county_demo, colour = "Democrat"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Party",  
                      labels = c("Democrat", "Republican")) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  labs(x = "Population Percentiles", 
       y = "Share of winning counties")
```

```{r}
boundaries <- c(0, 5000, 9000, 14000, 19000, 26000, 37000, 54000, 94000, 215000, Inf)

# Classify the counties into percentiles
data_population <- data %>% 
  mutate(pop_pctile = cut(total_population, breaks = boundaries, include.lowest = TRUE, labels = FALSE))

# data_population %>% 
#   filter(winning_party == "Republican") %>% 
#   ggplot(aes(x = pop_pctile)) +
#   geom_bar(aes(y = n()))
# 
# data_population %>% 
#   group_by(pop_pctile, winning_party) %>% 
#   summarise(n = n())
# 
# county_sum <- data_population %>%
#   filter(winning_party == "Republican") %>%
#   group_by(pop_pctile) %>%
#   summarize(count = n())

# Plot the data
# ggplot(county_sum, aes(x = pop_pctile, y = count)) +
#   geom_bar(stat = "identity")
```

From four graphs in @fig-dpc-income, we can conclude that, in general, the higher the DPC rate, the lower the support for Trump but higher for Biden. Besides, the Democrat is in favor of richer voters, where as Republican is in favor of low income counties. It is direct to ask whether these two variables have interaction effect on the voting patterns? However, it will be difficult to visualize the interaction effect between two continuous factors. Instead, I will manually set a new categorical variable `high_income` to denote whether the county has a high income with certain cutoff. Then the difference of correlations of the two groups as death per case rate changes. I will set different cutoffs to fully test the interaction effect.

```{r, fig.width=10, fig.height=4.5}
#| echo: false
#| message: false
#| label: fig-income_state
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: Variation in COVID-19 Mortality Rates Across States, Categorized by Income Levels and Dominant Political Preferences

# Find the wining data for each state
data_win_party <- data %>%
  group_by(state) %>%
  summarise(dem_votes = sum(vote_demo),
            rep_votes = sum(vote_rep), 
            .groups = 'drop') %>% 
  mutate(winning_party_state = ifelse(dem_votes > rep_votes, 'Democratic', 'Republican')) %>% 
  dplyr::select(state, winning_party_state)
# Calculate the average income
average_income_by_state <- data %>%
  group_by(state) %>%
  summarise(average_income = mean(mean_household_income), .groups = 'drop')
# Join the winning party and average income back to the original data
data_win_party_income <- data %>%
  left_join(data_win_party, by = "state") %>%
  left_join(average_income_by_state, by = "state")
# Arrange states by average income
ordered_states <- data_win_party_income %>%
  arrange(average_income) %>%
  .$state %>%
  unique()
# Create a factor with levels based on income-ordered states for plotting
data_win_party_income$state <- factor(data_win_party_income$state, levels = ordered_states)

average_dpc <- mean(data_win_party_income$mortrate)

ggplot(data_win_party_income, aes(x = state, y = mortrate, fill = winning_party_state)) +
  geom_boxplot() +
  geom_hline(yintercept = average_dpc, linetype = "dashed", color = "black") +
  annotate("text", x = Inf, y = average_dpc, label = " National Average", hjust = 1, vjust = -0.5, color = "black", size = 3) +
  scale_fill_manual(values = c('Democratic' = 'blue', 'Republican' = 'red')) +
  labs(y = 'Death Rate', x = 'State (Ascending Income)', 
       fill = "Winning Party") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

@@

The objective in this paper is to conduct the causal inference between COVID-19 and Trump's loss during 2020, the main methods implemented in this study will be the Propensity Score Matching (PSM). After find the causal inference, this paper will also conduct a counterfactual analysis to see whether Trump can re-elect if there was no COVID.

## Treatment

The treatment refers to the intervention or exposure that is being studied to understand its potential impact on the outcome. In other words, we implement certain interventions to a group of people but not for the rest. The group of people receiving the treatment is called the treatment group and control group for the rest. For instance, Jochen \[citation\] examined whether a new medical treatment called clampless off-pump coronary artery bypass (clampless OPCAB) will reduce the in-hospital mortality rate, compared to the traditional way. In their studies, the treatment is the new medical surgery and patients who took this surgery is the treatment group. They compared the mortality rates between the two groups and concluded that the new treatment can significantly lower the stroke and mortality rate.

Normally, the treatment should be established before conduct the analysis, this aims to reduce the risk of p-hacking. However, the objective of this paper is to find whether Trump lost the last Election due to COVID-19, it may not be possible to state the treatment in advance as all the counties in US had experienced COVI-19. Therefore this paper will firstly perform exploratory analysis to firstly find the treatment group which has the most significant difference in voting for Trump. This paper will set a cutoff defining whether a county has a high or low death per case rate. In addition, from @fig-dpc-income, low and high income counties also seem to have different voting behaviors. Therefore the treatment in this paper will incorporate whether a county has a high DPC and a high income levels with certain cutoffs.
